Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-14 paper/ResponseToReviews.tex

In addition to adding Figure 4-Figure supplement 4, we have added the following text related to this point:

\begin{quote}
\textsl{
One possible source of heterogeneity in the amount of viral mRNA per cell is variability in the timing of infection.
If some cells are infected earlier in the experiment than others, then they might have substantially more viral mRNA.
However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (Figure 4E, Figure 4-Figure~supplement~3), despite the fact that the potential time window for infection is much longer in the former sample.
Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C.
As shown in Figure 4-Figure~supplement 4, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since Figure 4B does not show the pervasive presence of mixed barcodes that would occur in this case.
Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.
}
\end{quote}
}

- Another potential contributor is the cell cycle status of the infected cells. Did the host transcript data shed light on whether cell cycle status influenced viral transcription levels? This is especially important to address given that some of the genes showing association with the viral burden are cell-cycle related.

{\color{black}
We have expanded the analysis to specifically look at cell-cycle related genes.
A gene-set analysis (Figure 9-Figure source data 2) finds that the G2-M cell-cycle checkpoint is associated with viral mRNA, with the expression of genes in this pathway typically expressed at lower levels in cells with more viral mRNA.
This analysis is now discussed in the second-to-last paragraph of the Results.
}

@@

The reviewers are correct that our statement about the role of co-infection in influencing cell-to-cell variation was too strong.
We have clarified this statement.
We have also added a figure supplement (Figure 4-Figure supplement 5) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
The new text reads:
\begin{quote}
\textsl{
Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 5).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 5).
}
\end{quote}
}

3. The analysis of host cell transcripts positively or negatively associated with high viral mRNA expression is pretty minimal. Do the host genes identified here match up with the results of previous studies that screened for host pro- and anti-viral factors (reviewed in Watanabe et al. DOI: 10.1016/j.chom.2010.05.008)? Also, targeted gene knockdown or over-expression experiments could help establish the causality underlying these relationships.

{\color{black}
We have substantially expanded the last section of the Results that focuses on cellular gene expression.
We call out the gene-set enrichments, which show that genes involved in oxidative stress responses are at higher levels in cells with more viral mRNA.
The possible pro-viral role of transient oxidative stress has been reported before, and we have performed new experiments (Figure 9-Figure supplement 2) that directly support it is pro-viral, suggesting causality.

We also compare the genes that we find significantly associated with viral mRNA to the 128 genes highlighted by Watanabe et al (2010) as having been found multiple times in genome-wide studies.
There is no overlap.
This fact is discussed in the last paragraph of the Results.
We theorize that this is because most of the cell-to-cell heterogeneity during infection of ``normal'' (e.g., not genetically perturbed) cells is \emph{not} due to changes in expression of the type of large-effect genes that are typically identified in genome-wide screens.
Rather, our results would seem to suggest that a substantial amount of the heterogeneity is due to viral / stochastic factors (e.g., failure to express certain viral genes), or perhaps more subtle changes in cellular state such as transient oxidative stress.

Importantly, we do \emph{not} believe that the main value of our approach is as a sort of single-cell gene knockdown screen to identify pro- or anti-viral genes.
Our goal is to figure out the actual extent and sources of cell-to-cell heterogeneity, which may not be due to primarily alterations in the expression level of single host genes with large effects.
}

### 2018-01-14 paper/paper.tex
added material about cell-cycle and more analysis of cellular genes

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell,hagai2017gene}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

@@

\subsection{Some host genes co-vary with viral gene expression.}
We examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
We identified 43 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
\jdbcomment{add gray arrow wedge}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.csv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

Many of the genes with increased expression in cells with more viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}). 
We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export~\citep{Amatore:2014cs}.
The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
%Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.

@@

For the analyses of cellular genes that differed in expression as a function of the amount of viral mRNA, we only considered cells that expressed all 8 viral mRNAs to avoid effects driven simply by viral gene absence.
We also only considered cellular genes in the differential gene analysis, since viral gene expression will tautologically co-vary with the amount of viral mRNA.
Additionally, because influenza virus has the capacity to degrade or prevent the synthesis of host mRNAs~\citep{BercovichKinori:2016iw} and contributes significantly to the total number UMIs in some cells, we calculate size factors (a scalar value representing efficiency of UMI capture) based on cellular transcripts alone. 
Finally, we assigned all cells a ceiling fraction of mRNA from virus of 25\% so that a few extremely high-expressing cells did not dominate. 
Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation. 
Figure~\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.

>>>>

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell,hagai2017gene}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

@@

\subsection{Some host genes co-vary with viral gene expression.}
We examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
We identified 43 cellular genes that co-varied with viral mRNA expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}, Figure~\ref{fig:cellulargenes}-source~data~\ref{figdata:cellulargenes}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
\jdbcomment{add gray arrow wedge}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}

\figdata{The full results of the differential expression test are in \texttt{p\_sig\_cellular\_genes.csv}. \label{figdata:cellulargenes}}

\figdata{A gene-set analysis for pathways associated with the amount of viral mRNA is in \texttt{p\_pathway\_enrichment.csv}. \label{figdata:geneset}}

\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}

\figsupp[Pre-treating to induce oxidative stress increases the fraction of infected cells expressing detectable viral protein. \label{figsupp:oxidative}]
{
\jdbcomment{Alistair, can you add figure and caption showing the increase in the percent of flu positive cells after treatment. It should just be possible to put the methods for this experiment in the figure caption.}
}
{\jdbcomment{add figure}}

A gene-set analysis shows that many cellular genes that are associated with the amount of viral mRNA are involved in the response to reactive oxygen species or hypoxia (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figdata:geneset}). 
Genes known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress are often expressed at higher levels in cells with more viral mRNA (Figure~\ref{fig:cellulargenes}).
These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}). 
We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export~\citep{Amatore:2014cs}.
The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
To directly test the effect of transient oxidative stress, we compared the fraction of cells that expressed detectable viral protein when infected either with or without pre-treatment to induce oxidative stress.
Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:oxidative} shows that the cells pre-treated to induce oxidative stress exhibited more frequent detectable expression of viral protein.
These results, in conjunction with the differential expression test in Figure~\ref{fig:cellulargenes} and the prior work mentioned above, suggest that oxidative stress acts in a proviral fashion.

The gene-set analysis also found that the amount of viral mRNA was associated with the expression of genes involved in the G2-M cell-cycle checkpoint (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figdata:geneset}).
The cell-cycle associated genes CCND3, MKI67, UBE2S, and CENPF are all expressed at significantly lower levels in cells with more viral mRNA (Figure~\ref{fig:cellulargenes}).
However, our data are not sufficient to determine whether the lower expression of these genes is a cause or effect of the reduction in viral mRNA.

Interestingly, none of the cellular genes that are significantly associated with the amount of viral mRNA in our study are among the 128 genes that \citet{watanabe2010cellular} report as having been identified multiple times in genome-wide screens for factors affecting influenza virus replication.
One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral factors, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.

@@

For the analyses of cellular genes that differed in expression as a function of the amount of viral mRNA, we only considered cells that expressed all 8 viral mRNAs to avoid effects driven simply by viral gene absence.
We also only considered cellular genes in the differential gene analysis, since viral gene expression will tautologically co-vary with the amount of viral mRNA.
Additionally, because influenza virus has the capacity to degrade or prevent the synthesis of host mRNAs~\citep{BercovichKinori:2016iw} and contributes significantly to the total number UMIs in some cells, we calculate size factors (a scalar value representing efficiency of UMI capture) based on cellular transcripts alone. 
Finally, we assigned all cells a ceiling fraction of mRNA from virus of 25\% so that a few extremely high-expressing cells did not dominate. 
Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation. 
Figure~\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
We performed the gene set analysis using the P -alues from the Monocle differentialGeneTest with \texttt{piano}~\citep{varemo2013enriching} using the hallmark gene set from GSEA v6~\citep{subramanian2005gene} and Fisher's method.
